The Elusive Benefits of Vagueness \protect \\ Evidence from Experiments

### 2018-08-02 springer/submission/v-book-2017.tex
massive restructuring Aug 2018

\documentclass[%
%			% choose man or doc immediately below
%doc,	% doc is single spaced
man,		% man is double spaced
floatsintext,%
apacite%
]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}

%\renewcommand{\baselinestretch}{1.7} % added 21 Sept 2016, copied from CRM book, but see documentclass option 'man' for "manuscript'' above which makes it double spaced

% MG change p's to 3 dp

\title{The Elusive Benefits of Vagueness \protect \\ Evidence from Experiments}
\date{\today}
%\authornote{latex compiled \today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, yet the benefits of vagueness for hearers and readers are proving to be elusive. We discuss a range of earlier controlled experiments with human participants, and we report on a new series of experiments that we conducted in recent years. These experiments, which focus on vague expressions that are part of referential noun phrases, aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. Having presented the evidence, we argue that the evidence supports a view where the benefits that vague terms exert are due to other influences, rather than to vagueness itself.}

@@

%Vagueness pervades the language that we use on a daily basis, and the challenge of understanding vague language has been a prominent concern in many areas of logic and linguistics, involving both theoretical and applied work, including the area known as Natural Language Generation (NLG).

%NLG systems can be thought of a small and imperfect models of human language use, and as such they have the ability to make questions about language very tangible, so let us take the perspective of an NLG system for a moment. In a nutshell, NLG systems transform data and formulas into language \cite<e.g.,>[]{reiter2000building}, making decisions between different formulations of the same information. For example, if the temperature is 27.2 degrees Celsius, the system could express this as ``27.2 degrees'', ``approximately 27 degrees', ``above 25 degrees", or ``warm'', and the system must decide between these.  The problem is important in the generation of textual weather reports from numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurements % such as oxygen saturation, heart rhythm, etc.
%\cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. Such systems often make decisions concerning the level of precision in the utterances that they generate on the basis of little more than intuition. Even when NLG systems are designed to mimic human language use \cite<e.g.,>{konstas2013global} there is no guarantee that these decisions taken by these systems benefit readers. A better understanding of the benefits (for readers and hearers) of different precision levels would allow these systems to become more useful. %The present article investigates the benefits, or de-benefits, of vagueness.

%. In the context of an NLG system faced with a practical decision of this kind, Lipman's question becomes ``Under what circumstances should vague terms be produced?''  Relevant applications include weather forecasting

%Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.}

@@

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions \cite{Lipman:2000fk, lipmanvague}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. In such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs. %The relevant conditions are, broadly speaking, that both interlocutors know all the relevant facts (e.g., both know the person's height precisely) and that the setting for the communication is co-operative. These conditions exclude deliberate deception, and rhetorical situations like political debate and advertising where the intention is to persuade the interlocutor to adopt some point of view, where the persuasion might not be intended to be in the addressee's best interest.

Lipman argued that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be. %, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.
Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, vanDeemterBook}. Prominent among these answers is the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this as the \emph{cost reduction} hypothesis. %The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). 
%This hypothesis predicts that when a vague expression can perform the same communicative task as a precise one, then, the things being equal, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

%The cost reduction hypothesis is of direct relevance to psycholinguists interested in language comprehension, and additionally to psycholinguists interested in language production, for example in connection with the question of audience design \cite{Clark1982287}. For, to the extent that speakers and writers choose vague expressions over and above crisp ones because the former are easier to process for hearers than the latter, the cost reduction hypothesis suggests that speakers design their utterance for optimal benefit to their hearers -- out of altruism, so to speak.

@@

\subsection{Our Approach to the Problem}\label{method}
%The experiments reported in the present paper put the cost reduction hypothesis to the test. The question that we are trying to answer is whether

@@

% KvD Feb 2017 (to address a criticism that came up in the reviews for the special issue):
We have chosen a narrative strategy in which we address a sequence of four experiments with human reader chronologically, explaining how each experiment helped us refine our research question. In order to do justice to our findings, we need to describe these experiments in a fair amount of detail.

@@

% MG General methodology moved to here: per-experiment method sections only talk about deviations from this general method
All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.
All $p$ values reported for linear models were calculated using the R package \emph{lmerTest} \cite{lmerTest}. The complete data and analysis of our first experiment are available from 
{\tt http://homepages.abdn.ac.uk/k.vdeemter/pages/vagueness-data-analysis/e1.html}, and analogous for experiments 2-4.\\[2ex]

% KvD I've moved the following two paragraphs towards the end of the section.
% MG added/moved out of expt 1

\noindent When the distance grows between two numbers, they become more easily distinguishable: the \emph{numerical distance effect} has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene1996}. 
We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. 
% MG added/moved out of expt 1
When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. 
We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

% KvD added the following from a previous version of the paper:
There is abundant evidence \cite<e.g.,>{trick1994small} that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment \cite{our-earlier-paper} in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper. We found that, when participants were confronted with % MG changed the next line and added the footnote to remove the doubts we had about whether a dot array must have regular layout, or can have irregular layout - I'm still not sure this is a good description though
a stimulus consisting of two squares containing different numbers of dots\footnote{Such a stimulus is referred to hereafter as consisting of a set of {\em dot arrays}. The number of dots in an array is referred to as its cardinality. The physical arrangement of dots in each array is irregular.}, 
instructions of the form \emph{Choose the square with \emph{n} dots} led to consistently faster response times than instructions of the form \emph{Choose the square with many/few dots} when $2 \leq n \leq 5$; the converse was true for $n>5$. Given these findings, we henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

%%%%%%%%%%%%%%%%%%%%%%%%
%                            E 1                                     %
%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\subsubsection{Response times}
{\bf Response times} (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants. Analysis confirms hypothesis H1:

@@

% KvD: I've added some syntactic sugar, in a number of places, to clarify the relation between Ex 1 and Ex 2.

@@

% KvD I've re-instated the table below, because this makes it easier to understand the instruction format/selection algorithm confound.

@@

%\subsubsection{Response times} % E2
% MG added the values for these stats on 26 Sept 2016
Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:
{\small
\begin{APAenumerate}
	\item [(H1)] The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against Hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06$, $se=.01$, $t=4.6$, $p<0.0001$). 
	\item [(H2)] In focussed comparisons, vagueness was significantly disadvantageous in the \emph{numeric} conditions ($\beta=0.09$, $se=0.02$, $t=4.5$, $p<0.0001$), and non-significantly disadvantageous in the \emph{verbal} conditions ($\beta=0.02$, $se=0.02$, $t=1.5$, $p=0.1550$), offering partial evidence against Hypothesis 2. The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to a significant overall interaction effect between vagueness and instruction format ($\beta=-0.13$, $se=0.02$, $t=-6.6$, $p=0.0169$).
	\item [(H3)] There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1, but suggesting that instruction format rather than vagueness drove the effect we observed in Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; ($\beta=-0.36$, $se=0.07$, $t=-5.1$, $p<0.0001$).
	\item [(H4)] There was a significant main effect of item (i.e., of which triple of numbers of dots were used in the stimulus): ($\beta=0.12$, $se=0.02$, $t=7.1$, $p<0.0001$). This effect seems likely to be due to the very fast responses for stimuli using arrangements of the smallest numbers of dots (6,15,24), which also had the largest difference in ratio of smallest number to largest number in the stimulus, suggesting that stimuli using these numbers of dots may have been particularly discriminable for participants.
\end{APAenumerate}
}

%\subsubsection{Borderline responses} % E2
% MG added the values for these stats on 26 Sept 2016
A generalized linear mixed model \cite{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}. Participants chose the borderline square on $16.6\%$ of trials overall.
{\small 
\begin{APAenumerate}
	\item [(H5)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{APAenumerate}
}

@@

% KvD I've expanded the following:
However, once again there is a possibly confounding factor. Observe that, in Experiment 2, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: To see this, consider the task of selecting the dot array that contains ``few dots": to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains "16 dots" seems to require the participant to estimate, and then {\em match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm".% (which is known to be faster than matching). % KvD I think we need to say more about this, citing some appropriate papers to support this idea. \textcolor{red}{{\bf--Insert citation}}

@@

%KvD I think we should motivate our choice of materials (especially "far ..er")
%\textcolor{red}{{\bf KvD I think we should motivate our choice of materials (especially "far ..er")}}
% DONE 28 Sept 2017

@@

\begin{figure}[htbp]
\centering
\fitfigure{images/e3-rtplot-1.pdf}
%\includegraphics[trim = 20mm 0mm 20mm 0mm, clip, width=.75\textwidth]{images/e3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 3 where all instructions were numeric}
\label{resultse3}
\end{figure}

% MG added the values for these stats on 26 Sept 2016
{\small
\begin{APAenumerate}
	\item [(H1)] Vague instructions were non-significantly easier for the reader than crisp instructions when vagueness was considered as a main effect ($\beta=-0.01$, $se=0.01$, $t=-0.4$, $p=0.7158$). 
	\item [(H2)] There was a statistically significant main effect of selection task, with the comparison task speeding responses compared to the matching task ($\beta=0.16$, $se=0.03$, $t=6.2$, $p<0.0001$). 
	\item [(H3)] Vagueness exerted effects in different directions for the comparison task and for the matching task: there was a significant interaction between vagueness and selection task ($\beta=0.13$, $se=0.03$, $t=4.2$, $p<0.0001$). 
Separate analyses were conducted testing for effects of vagueness at each level of the selection task.
Within the comparison task vagueness significantly speeded response times compared with crisp controls ($\beta=-0.07$, $se=0.02$, $t=-3.5$, $p<0.0012$). 
Within the matching task vagueness significantly \emph{slowed} response times compared with crisp controls ($\beta=0.06$, $se=0.02$, $t=2.9$, $p<0.0061$). 
\end{APAenumerate}
}
% MG edited this 27 Sept 2016:
%The cost reduction account was wrong to predict main effect advantages for vagueness, and wrong to predict that vagueness should be beneficial at each level of the selection task: however vagueness was advantageous in the comparison task.
In other words, the cost reduction account was wrong to predict significant main effect advantages for vagueness (although there was a non-significant trend in the \emph{direction} predicted by the cost reduction account), and wrong to predict that vagueness should be beneficial at each level of the selection task: however vagueness was significantly advantageous in the comparison task.

@@

This experiment mirrors Experiment 3, but focussing on instructions that did not use a number. 
We manipulated vagueness and the selection task (comparison and matching). 
In order to implement the experiment without mentioning numbers in the instructions, we changed the sequence of each trial to include a `target' (i.e., a dot array of a particular cardinality) before the instruction, so that we could then refer back to the target's cardinality in the instruction using expressions like \emph{the same number of dots as the target}; \emph{fewer dots than the target}.
%
This presentation of a target before the main body of the trial shares some features with \citeA[Experiment 2]{Izard20081221}, although in that experiment participants were told the cardinality of the target (called an \emph{inducer} in that paper) whereas in our experiment we did not tell participants the cardinality of the prime array.
An item was thus a combination of a target dot array, an instruction that did not contain a number, and a set of dot arrays taking their cardinalities from the same triples used in Experiment 2.
Table \ref{Instructions for e4} spells out how the instructions were constrained not to mention a numeral and gives examples of targets.

@@

% MG added the values for these stats on 27 Sept 2016
{\small
\begin{APAenumerate}
	\item [(H1)] There was no significant main effect of vagueness ($\beta=-0.02$, $se=0.01$, $t=-1.5$, $p=0.1296$). 
	\item [(H2)] There was a main effect of selection, with comparison task instructions leading to faster responses than the matching task instructions ($\beta=0.18$, $se=0.02$, $t=10.4$, $p<0.0001$).  This effect was in the same direction as Experiment 3. 
	\item[(H3)] Vagueness did exert different effects depending on the selection task (main interaction effect of vagueness by selection $\beta=0.12$, $se=0.02$, $t=5.1$, $p<0.0001$). 
Separate analyses of the effect of vagueness were conducted for the comparison task and for the matching task using Bonferroni-adjusted significance thresholds. 
In the comparison task, vagueness resulted in faster response times ($\beta=-0.08$, $se=0.02$, $t=-4.3$, $p<0.0001$). 
In the matching task vagueness slowed response times ($\beta=0.05$, $se=0.01$, $t=3.7$, $p=0.0004$). 
These results are in the same direction as Experiment 3.
\end{APAenumerate}
}

@@

To summarise our findings, Experiment 1 showed that responses were faster and more accurate when the instructions were vague than when they were crisp, but the experiment could not distinguish effects of vagueness from those of number-avoidance or selection task: the vague conditions were also in verbal rather than numerical format; and mandated a comparison strategy rather than a matching strategy.  Experiment 2 showed that number avoidance in the verbal format instructions is an important factor driving the faster response times in the task, and that vagueness does not have any additional benefit in either the verbal format instructions or the numerical format instructions.

However, Experiment 2 did not distinguish benefits of number avoidance from benefits of the comparison selection task. In Experiments 3 and 4 we manipulated vagueness and the selection task, separately at each level of numerical format. Across the two experiments, we found that the comparison-task instructions attracted faster response times than the matching-task instructions. Within the two experiments we found that vagueness exerts benefits when the selection task is \emph{comparison}, but not when the task is \emph{matching}.

>>>>

\documentclass[%
%			% choose man or doc immediately below
%doc,	    % doc is single spaced
man,		% man is double spaced
floatsintext,%
apacite%
]{apa6} % I'm using a local (as in, in this folder) apa6.cls which edits footnotesep from 16pt which is the real apa6 in 2018 back to 6pt in order to make it line up with the pdf generated for the 2017 submission
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}

\title{The Elusive Benefits of Vagueness \protect \\ Evidence from Experiments}
\date{\today}
\authornote{This is a layout-identical version of the document submitted for review in May 2017. LaTeX compiled \today~at \currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, even in situations where vagueness could have been avoided (i.e., where vagueness is used "strategically"). Yet the benefits of vagueness for hearers and readers are proving to be elusive. We discuss a range of earlier controlled experiments with human participants, and we report on a new series of experiments that we conducted in recent years. These experiments, which focus on vague expressions that are part of referential noun phrases, aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. Having presented the evidence, we argue that the evidence supports a view where the benefits that vague terms exert are due to other influences, and not to vagueness
itself.}

@@

@@

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions \cite{Lipman:2000fk, lipmanvague}, which are
often used even in situations where the speaker could have used an expression that is not vague (i.e., crisp); in these situations we say that vagueness is used \emph{strategically}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. In such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs.

Lipman argued that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be. Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, vanDeemterBook}. One of the most promising answers appears to be the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this as the \emph{cost reduction} hypothesis.

This article brings an experimental approach to these issues, focussing on vagueness in descriptions (e.g., "the square with few dots") and its effect on the hearer's ability to act on a given description, as measured by the time that it takes hearers to click on the referent of a description\footnote{Other metrics could have been chosen, such as hearers' ability to remember information, for example, or error rates. Although error rates play a minor role in the present paper, for reasons that will become clear, we focus on response times in particular.}. We find that, although hearer benefits from vague descriptions are straightforward to demonstrate in many cases, a closer experimental analysis militates against the conclusion that vagueness itself -- as defined above, in terms of the existence of borderline cases -- lies at the heart of these results. Instead, it is other factors, such as the presence of an overt numerical expression in the description, that proved to be decisive. We believe that, despite the fact that our experiments are unavoidably focussed only on a specific class of vague expressions (since any experiment can only deal with a limited number of different stimuli), these findings are potentially important, because they call into question whether "strategic" vagueness (i.e., vagueness where the speaker had a choice, because she could have been produced a crisp expression instead) has any advantages at all. In other words, returning to Lipman's question, it is possible that vagueness has evolved partly as a necessary evil (e.g., because of the limits of observation and prediction) and partly as a side effect of other factors.

\section{Related work}

@@

\subsection{Our Approach to the Problem}\label{method}

@@

We have chosen a narrative strategy in which we address a sequence of four experiments with human reader chronologically, explaining how each experiment helped us refine our research question. In order to do justice to our findings, we need to describe these experiments in a fair amount of detail.

@@

All the experiments shared the following properties: 
Stimuli were created using the language GNU Octave \cite{eaton:2002} and the Psychophysics Toolbox extensions \cite{ptbx1, ptbx2}.
The position of the dots was randomised per-trial.
The order in which trials were presented was randomised per-participant. 
There were 256 trials, presented in 4 blocks of 64 each, between which the participant could rest. 
A MacBook Pro laptop computer with a 13 inch screen presented the stimuli to the participants and recorded responses. 
Participants were recruited using email lists at the University of Aberdeen, and paid ten pounds for participating.
All participants self-reported fluency in English, and had normal, or corrected-to-normal vision.
The experiment was conducted in a quiet room. 
Participants were asked to respond as quickly as possible while avoiding errors.
There was a block of practice trials after which participants could ask any questions, following which the experimenter left the room.
All $p$ values reported for linear models were calculated using the R package \emph{lmerTest} \cite{lmerTest}. The complete data and analysis of our first experiment are available from 
{\tt http://homepages.abdn.ac.uk/k.vdeemter/pages/vagueness-data-analysis/e1.html}, and analogous for experiments 2-4.\\[2ex]

\noindent When the distance grows between two numbers, they become more easily distinguishable: the \emph{numerical distance effect} has been shown for comparing the cardinality of two sets of dots \cite{van123} and for processing Arabic numerals and number words \cite{Dehaene1996}. We manipulated the number of dots in each array such that some sets of arrays had smaller numerical distances and others had larger numerical distances. Where a number was mentioned in the instructions, it was always in the form of an Arabic numeral. When two numbers are presented with the smaller on the left, this left-side presentation facilitates responses indicating the smaller number: the \emph{Spatial-Numerical Association of Response Codes (SNARC)} effect \cite{dehaene1993mental, gevers2006automatic}. We controlled which side the smaller number appeared on to avoid systematic influences of this effect.

There is abundant evidence \cite<e.g.,>{trick1994small} that very small (i.e., \emph{subitizable}) quantities are recognised and processed by a distinct psychological mechanism that differs from that used to process larger quantities. We performed a pilot experiment \cite{our-earlier-paper} in which we were able to confirm this finding in the experimental settings on which we are focussing in this paper. We found that, when participants were confronted with a stimulus consisting of two squares containing different numbers of dots\footnote{Such a stimulus is referred to hereafter as consisting of a set of {\em dot arrays}. The number of dots in an array is referred to as its cardinality. The physical arrangement of dots in each array is irregular.}, instructions of the form \emph{Choose the square with \emph{n} dots} led to consistently faster response times than instructions of the form \emph{Choose the square with many/few dots} when $2 \leq n \leq 5$; the converse was true for $n>5$. Given these findings, we henceforth focussed our studies on non-subitizable numbers, because it is there that vagueness is expected to have benefits.

@@

{\bf Response times} (RTs) for trials with erroneous responses were discarded, leading to the loss of 354 trials from 5120, representing 6.9\% of the trials. 
The correct response RTs were trimmed at 2.5 standard deviations for each subject, leading to the loss of a further 160 trials, or 3.4\% of the remaining correct responses. 
Means for response times and error rates are given in Fig. (\ref{resultse1}).
A linear mixed model of RT was built, with vagueness and numerical distance and their interaction as independent variables, and with random slopes for vagueness and numerical distance over participants. Analysis confirms hypothesis H1:

@@

@@

@@

%\subsubsection{Response times} % E2
Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:
{\small
\begin{APAenumerate}
	\item [(H1)] The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against Hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06$, $se=.01$, $t=4.6$, $p<0.0001$). 
	\item [(H2)] In focussed comparisons, vagueness was significantly disadvantageous in the \emph{numeric} conditions ($\beta=0.09$, $se=0.02$, $t=4.5$, $p<0.0001$), and non-significantly disadvantageous in the \emph{verbal} conditions ($\beta=0.02$, $se=0.02$, $t=1.5$, $p=0.1550$), offering partial evidence against Hypothesis 2. The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to a significant overall interaction effect between vagueness and instruction format ($\beta=-0.13$, $se=0.02$, $t=-6.6$, $p=0.0169$).
	\item [(H3)] There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1, but suggesting that instruction format rather than vagueness drove the effect we observed in Experiment 1 (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; ($\beta=-0.36$, $se=0.07$, $t=-5.1$, $p<0.0001$).
	\item [(H4)] There was a significant main effect of item (i.e., of which triple of numbers of dots were used in the stimulus): ($\beta=0.12$, $se=0.02$, $t=7.1$, $p<0.0001$). This effect seems likely to be due to the very fast responses for stimuli using arrangements of the smallest numbers of dots (6,15,24), which also had the largest difference in ratio of smallest number to largest number in the stimulus, suggesting that stimuli using these numbers of dots may have been particularly discriminable for participants.
\end{APAenumerate}
}

A generalized linear mixed model \cite{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}. Participants chose the borderline square on $16.6\%$ of trials overall.
{\small 
\begin{APAenumerate}
	\item [(H5)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{APAenumerate}
}

@@

However, once again there is a possibly confounding factor. Observe that, in Experiment 2, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: To see this, consider the task of selecting the dot array that contains ``few dots": to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains "16 dots" seems to require the participant to estimate, and then {\em match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm".

@@

@@

\begin{figure}[htbp]
\centering
\fitfigure{images/e3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 3 where all instructions were numeric}
\label{resultse3}
\end{figure}

{\small
\begin{APAenumerate}
	\item [(H1)] Vague instructions were non-significantly easier for the reader than crisp instructions when vagueness was considered as a main effect ($\beta=-0.01$, $se=0.01$, $t=-0.4$, $p=0.7158$). 
	\item [(H2)] There was a statistically significant main effect of selection task, with the comparison task speeding responses compared to the matching task ($\beta=0.16$, $se=0.03$, $t=6.2$, $p<0.0001$). 
	\item [(H3)] Vagueness exerted effects in different directions for the comparison task and for the matching task: there was a significant interaction between vagueness and selection task ($\beta=0.13$, $se=0.03$, $t=4.2$, $p<0.0001$). 
Separate analyses were conducted testing for effects of vagueness at each level of the selection task.
Within the comparison task vagueness significantly speeded response times compared with crisp controls ($\beta=-0.07$, $se=0.02$, $t=-3.5$, $p<0.0012$). 
Within the matching task vagueness significantly \emph{slowed} response times compared with crisp controls ($\beta=0.06$, $se=0.02$, $t=2.9$, $p<0.0061$). 
\end{APAenumerate}
}
In other words, the cost reduction account was wrong to predict significant main effect advantages for vagueness (although there was a non-significant trend in the \emph{direction} predicted by the cost reduction account), and wrong to predict that vagueness should be beneficial at each level of the selection task: however vagueness was significantly advantageous in the comparison task.

@@

This experiment mirrors Experiment 3, but focussing on instructions that did not use a number. 
We manipulated vagueness and the selection task (comparison and matching). 
In order to implement the experiment without mentioning numbers in the instructions, we changed the sequence of each trial to include a `target' (i.e., a dot array of a particular cardinality) before the instruction, so that we could then refer back to the target's cardinality in the instruction using expressions like \emph{the same number of dots as the target}; \emph{fewer dots than the target}.
This presentation of a target before the main body of the trial shares some features with \citeA[Experiment 2]{Izard20081221}, although in that experiment participants were told the cardinality of the target (called an \emph{inducer} in that paper) whereas in our experiment we did not tell participants the cardinality of the prime array.
An item was thus a combination of a target dot array, an instruction that did not contain a number, and a set of dot arrays taking their cardinalities from the same triples used in Experiment 2.
Table \ref{Instructions for e4} spells out how the instructions were constrained not to mention a numeral and gives examples of targets.

@@

{\small
\begin{APAenumerate}
	\item [(H1)] There was no significant main effect of vagueness ($\beta=-0.02$, $se=0.01$, $t=-1.5$, $p=0.1296$). 
	\item [(H2)] There was a main effect of selection, with comparison task instructions leading to faster responses than the matching task instructions ($\beta=0.18$, $se=0.02$, $t=10.4$, $p<0.0001$).  This effect was in the same direction as Experiment 3. 
	\item[(H3)] Vagueness did exert different effects depending on the selection task (main interaction effect of vagueness by selection $\beta=0.12$, $se=0.02$, $t=5.1$, $p<0.0001$). 
Separate analyses of the effect of vagueness were conducted for the comparison task and for the matching task using Bonferroni-adjusted significance thresholds. 
In the comparison task, vagueness resulted in faster response times ($\beta=-0.08$, $se=0.02$, $t=-4.3$, $p<0.0001$). 
In the matching task vagueness slowed response times ($\beta=0.05$, $se=0.01$, $t=3.7$, $p=0.0004$). 
These results are in the same direction as Experiment 3.
\end{APAenumerate}
}

@@

To summarise our findings, we asked why strategic vagueness is as frequent as it is and we decided to focus on what we see as the most promising explanation, namely that vague expressions are easy to process by speakers and hearers: the cost hypothesis, as we have called it. We decided to test this hypothesis by experimentally investigating whether vague descriptions are resolved by hearers more quickly than crisp ones. Although we were able to find some interesting (and statistically highly significant) effects, it appears to us that if our sequence of experiments is assessed as a whole, it cannot be seen as confirmation of the cost hypothesis.
